cardinalR: Generating interesting high-dimensional data structures

### 2025-10-02 paper-cardinalR.Rmd
Added Swissroll section

>>>>

#### Polynomial

A polynomial structure generates data points that follow non-linear curvilinear relationships, such as quadratic or cubic trends, in \gD{} space. To extend these patterns into high-dimensional settings, additional noise dimensions can be added. These patterns are useful for evaluating how well algorithms capture smooth, non-linear trajectories and curvature in the data. We provide functions for generating quadratic and cubic structures, enabling controlled experiments with different degrees of polynomial complexity. Table `r knitr::asis_output(ifelse(knitr::is_html_output(), '\\@ref(tab:polynomial-tb-html)', '\\@ref(tab:polynomial-tb-pdf)'))` summarizes these functions and their purposes.

```{r}
polynomial_tb <- tibble(fun = c("gen_quadratic",
                                "gen_cubic"), 
                        exp = c("Quadratic pattern.",
                                "Cubic pattern."))
```

```{r polynomial-tb-html, eval=knitr::is_html_output()}
polynomial_tb |> 
  kable(caption = "cardinalR polynomial data generation functions", col.names = c("Function", "Explanation")) 
```

```{r polynomial-tb-pdf, eval=knitr::is_latex_output()}
polynomial_tb |> 
  kable(caption = "cardinalR polynomial data generation functions", format="latex", col.names = c("Function", "Explanation"), booktabs = T)  |>
  column_spec(1, width = "4cm") |>
  column_spec(2, width = "8cm")
```

The first is the quadratic curve of $n$ points in two dimensions. This is generated using `gen_quadratic(n, range)`. The independent variable is defined as $X_1 \sim U(\text{range}[1], \text{range}[2])$, and a raw polynomial basis of degree 2 is applied to form $X_2 = X_1 - X_1^2 + \varepsilon_2$, where $\varepsilon_2 \sim U(0, 0.5)$. This produces a smooth parabolic arc opening downward, with vertical jitter introduced by the noise term.

```{r data-quadratic, echo=TRUE}
quadratic <- gen_quadratic(n = 1000)
```

The second is the cubic curve of $n$ points in two dimensions. This is generated using `gen_cubic(n, range)`. The independent variable is defined as $X_1 \sim U(\text{range}[1], \text{range}[2])$, and a raw polynomial basis of degree $3$ is applied to construct $X_2 = X_1 + X_1^2 - X_1^3 + \varepsilon_2$, where $\varepsilon_2 \sim U(0, 0.5)$. This produces a more complex curvilinear structure than the quadratic case, with both upward and downward turning points.

```{r data-cubic, echo=TRUE}
cubic <- gen_cubic(n = 1000)
```

#### Swiss Roll

To further generalize the Swiss roll structure and introduce realistic noise, we define a function `gen_swissroll(n, w)`, where $n$ is the number of points, $p$ is the total number of dimensions, and $w$ is the vertical range in the third dimension (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:swissroll)"} else { "Figure \\@ref(fig:swissroll-proj)"})`). The first three dimensions form the classic $3\text{-}D$ Swiss roll shape. The $$X_1 = t \cos(t), \quad X_2 = t \sin(t), \quad X_3 \sim U(w_1, w_2), \text{ where } t \sim U(0, 3\pi).$$

```{r data-swissroll, echo=TRUE}
swissroll <- gen_swissroll(n = 1000, w = c(-1, 1))
```

```{r}
swissroll_lang <- langevitour::langevitour(swissroll, levelColors = "black", enableControls = FALSE, width = "400px", height = "300px")
```

```{r swissroll-proj1}
scaled_data <- scale_data_manual(swissroll)

swissroll_proj1 <- plot_proj(
  proj_obj = proj_obj1, 
  point_param = c(1.5, 0.2, "#000000"), # size, alpha, color
  plot_limits = c(-0.65, 0.7), 
  title = "a1", 
  cex = 2, 
  axis_text_size = 5,
  is_color = FALSE)

```{r swissroll-proj2}

swissroll_proj2 <- plot_proj(
  proj_obj = proj_obj2, 
  point_param = c(1.5, 0.2, "#000000"), # size, alpha, color
  plot_limits = c(-0.7, 0.65), 
  title = "a2", 
  cex = 2, 
  axis_text_size = 5,
  is_color = FALSE)

```{r swissroll-proj3}

proj_obj3 <- get_projection(projection = projection, 
                            proj_scale = 1.2, 
                            scaled_data = scaled_data, 
                            axis_param = list(limits = 0.6,
                                              axis_scaled = 1.5, 
                                              axis_pos_x = -0.55, 
                                              axis_pos_y = -0.55, 
                                              threshold = 0.016))

swissroll_proj3 <- plot_proj(
  proj_obj = proj_obj3, 
  point_param = c(1.5, 0.2, "#000000"), # size, alpha, color
  plot_limits = c(-0.65, 0.5), 
  title = "a3", 
  cex = 2, 
  axis_text_size = 5,
  is_color = FALSE)

```{r swissroll, eval=knitr::is_html_output(), fig.cap="`langevitour` output  of the `swissroll` data in $3\\text{-}D$."}

swissrollfig <- bscols(
  htmltools::div(
    style = "display: grid; grid-template-columns: 1fr 1fr;",
    swissroll_lang
  ),
  device = "xs"
)

class(swissrollfig) <- c(class(swissrollfig), "htmlwidget")

swissrollfig
```

```{r swissroll-proj, eval=knitr::is_latex_output(), fig.pos="H", fig.cap="Three $2\\text{-}D$ projections from $3\\text{-}D$, for the `swissroll` data."}
#| fig-width: 12
#| fig-height: 4

swissroll_proj1 + swissroll_proj2 + swissroll_proj3 +
  plot_layout(ncol = 3, guides = "collect") 
```

### 2025-10-02 paper-cardinalR.tex
Added Swissroll section

>>>>

\paragraph{Polynomial}\label{polynomial}

A polynomial structure generates data points that follow non-linear curvilinear relationships, such as quadratic or cubic trends, in $2\text{-}D${} space. To extend these patterns into high-dimensional settings, additional noise dimensions can be added. These patterns are useful for evaluating how well algorithms capture smooth, non-linear trajectories and curvature in the data. We provide functions for generating quadratic and cubic structures, enabling controlled experiments with different degrees of polynomial complexity. Table \ref{tab:polynomial-tb-pdf} summarizes these functions and their purposes.

\caption{\label{tab:polynomial-tb-pdf}cardinalR polynomial data generation functions}
\centering
\begin{tabular}[t]{>{\raggedright\arraybackslash}p{4cm}>{\raggedright\arraybackslash}p{8cm}}
\toprule
Function & Explanation\\
\midrule
gen\_quadratic & Quadratic pattern.\\
gen\_cubic & Cubic pattern.\\
\bottomrule
\end{tabular}
\end{table}

The first is the quadratic curve of \(n\) points in two dimensions. This is generated using \texttt{gen\_quadratic(n,\ range)}. The independent variable is defined as \(X_1 \sim U(\text{range}[1], \text{range}[2])\), and a raw polynomial basis of degree 2 is applied to form \(X_2 = X_1 - X_1^2 + \varepsilon_2\), where \(\varepsilon_2 \sim U(0, 0.5)\). This produces a smooth parabolic arc opening downward, with vertical jitter introduced by the noise term.

\begin{verbatim}
quadratic <- gen_quadratic(n = 1000)
\end{verbatim}

The second is the cubic curve of \(n\) points in two dimensions. This is generated using \texttt{gen\_cubic(n,\ range)}. The independent variable is defined as \(X_1 \sim U(\text{range}[1], \text{range}[2])\), and a raw polynomial basis of degree \(3\) is applied to construct \(X_2 = X_1 + X_1^2 - X_1^3 + \varepsilon_2\), where \(\varepsilon_2 \sim U(0, 0.5)\). This produces a more complex curvilinear structure than the quadratic case, with both upward and downward turning points.

\begin{verbatim}
cubic <- gen_cubic(n = 1000)
\end{verbatim}

\paragraph{Swiss Roll}\label{swiss-roll}

To further generalize the Swiss roll structure and introduce realistic noise, we define a function \texttt{gen\_swissroll(n,\ w)}, where \(n\) is the number of points, \(p\) is the total number of dimensions, and \(w\) is the vertical range in the third dimension (Figure \ref{fig:swissroll-proj}). The first three dimensions form the classic \(3\text{-}D\) Swiss roll shape. The \[X_1 = t \cos(t), \quad X_2 = t \sin(t), \quad X_3 \sim U(w_1, w_2), \text{ where } t \sim U(0, 3\pi).\]

\begin{verbatim}
swissroll <- gen_swissroll(n = 1000, w = c(-1, 1))
\end{verbatim}

\begin{figure}[H]
\includegraphics[width=1\linewidth]{paper-cardinalR_files/figure-latex/swissroll-proj-1} \caption{Three $2\text{-}D$ projections from $3\text{-}D$, for the `swissroll` data.}\label{fig:swissroll-proj}
\end{figure}
